---
title: "Import and representation of MERFISH mouse ileum data"
author: "Ludwig Geistlinger and Robert Gentleman"
affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse ileum}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r, message = FALSE}
library(ggplot2)
library(vroom)
library(jsonlite)
library(BumpyMatrix)
library(SpatialExperiment)
```

Start with downloading from and unzipping the dataset from:
https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.jm63xsjb2


```{r}
data.dir <- "data_release_baysor_merfish_gut"
raw.dir <- file.path(data.dir, "raw_data")
proc.dir <- file.path(data.dir, "data_analysis")
baysor.dir <- file.path(proc.dir, "baysor") 
cellpose.dir <- file.path(proc.dir, "cellpose") 
baysor.cellpose.dir <- file.path(proc.dir, "baysor_membrane_prior") 
```

# Data

Def. ileum: the final and longest segment of the small intestine.

## Raw data

What's there:

```{r}
list.files(raw.dir)
```

mRNA molecule data: 820k observations for 241 genes

```{r, message = FALSE}
mol.file <- file.path(raw.dir, "molecules.csv")
mol.dat <- vroom::vroom(mol.file)
mol.dat <- data.frame(mol.dat)
dim(mol.dat)
head(mol.dat)
length(unique(mol.dat$gene))
```

Image data: 

1. [DAPI](https://en.wikipedia.org/wiki/DAPI) stain signal:

```{r, fig.height = 10}
dapi.file <- file.path(raw.dir, "dapi_stack.tif")
dapi.img <- SpatialExperiment::SpatialImage(dapi.file)
dapi.img <- SpatialExperiment::mirrorImg(dapi.img, "h")
class(dapi.img)
SpatialExperiment::imgSource(dapi.img)
plot(SpatialExperiment::imgRaster(dapi.img))
```

2. Membrane Na+/K+ - ATPase immunofluorescence signal:

```{r, fig.height = 10}
mem.file <- file.path(raw.dir, "membrane_stack.tif")
mem.img <- SpatialExperiment::SpatialImage(mem.file)
mem.img <- SpatialExperiment::mirrorImg(mem.img, "h")
class(mem.img)
SpatialExperiment::imgSource(mem.img)
plot(SpatialExperiment::imgRaster(mem.img))
```

## Processed data

### Baysor (mRNA-only) segmentation

1. Cell-by-gene matrix. Rows: genes; Cols: cells

```{r, message = FALSE}
counts.file <- file.path(baysor.dir, "segmentation", "segmentation_counts.tsv")
counts.baysor <- vroom::vroom(counts.file, col_names = FALSE)
counts.baysor <- data.frame(counts.baysor)
genes <- counts.baysor[,1]
counts.baysor <- as.matrix(counts.baysor[,-1])
rownames(counts.baysor) <- genes
dim(counts.baysor)
counts.baysor[1:5,1:5]
```

2. Cell metadata. Each cell corresponds to each column in `segmentation_counts.csv`.

```{r, message = FALSE}
cdat.file <- file.path(baysor.dir, "segmentation", "segmentation_cell_stats.csv")
cdat.baysor <- vroom::vroom(cdat.file)
cdat.baysor <- data.frame(cdat.baysor)
colnames(counts.baysor) <- cdat.baysor$cell
dim(cdat.baysor)
head(cdat.baysor)
```

3. Segmentation and mRNA metadata. Each row corresponds to one mRNA in `raw_data/molecules.csv`

```{r, message = FALSE}
rdat.file <- file.path(baysor.dir, "segmentation", "segmentation.csv")
rdat.baysor <- vroom::vroom(rdat.file)
rdat.baysor <- data.frame(rdat.baysor)
rdat.baysor <- subset(rdat.baysor, cell != 0)
dim(rdat.baysor)
head(rdat.baysor)
```   

4. Polygon segmentation borders:

Segmentation borders are provided in JSON format, and
we have polygon segmentation borders for each of the nine z-layers:

```{r}
poly.file <- file.path(baysor.dir, "segmentation", "poly_per_z.json")
pdat <- jsonlite::fromJSON(poly.file)
dim(pdat)
colnames(pdat)
```

Now let's look at the segmentation borders of the first cell in the first z-layer:

```{r}
head(pdat[1,2][[1]]$coordinates[[1]])
pdf <- as.data.frame(pdat[1,2][[1]]$coordinates[[1]][1,,])
colnames(pdf) <- c("x", "y")
ggplot(pdf, aes(x = x, y = y)) + geom_polygon()
```

That is a terribly nested data structure, let's reshape that to a `data.frame`
so we can more conveniently work with that.

We therefore first write a function to pull out the coordinates for one z-layer
at a time:

```{r}
pcoords <- pdat[1,2][[1]]$coordinates
getSegmentationDF <- function(coords)
{
    y <- lapply(coords, function(x) as.data.frame(x[1,,])) 
    y <- lapply(y, as.matrix)
    z <- do.call(rbind, y)
    n <- vapply(y, nrow, numeric(1)) 
    colnames(z) <- c("x", "y")
    z <- cbind(z, cell = rep(seq_along(y), n))    
    return(z)
}
head(getSegmentationDF(pcoords))
```

We then apply the function to each z-layer to arrive at an overall `data.frame`:

```{r}
dfl <- lapply(pdat[,2], function(x) getSegmentationDF(x$coordinates))
n <- vapply(dfl, nrow, numeric(1))
df <- do.call(rbind, dfl)
df <- cbind(df, z = rep(seq_along(dfl), n))
df <- data.frame(df)
head(df)
dim(df) 
```

Ok, now we can look into how many cells / polygons do we have for each z-layer:

```{r}
spl <- split(df$cell, df$z)
spl <- lapply(spl, unique)
lengths(spl)
```

Question: how to relate that to the 5,800 cells that we have in the segmentation 
counts matrix? 

Somehow the numbers don't add up with what we have in the segmentation table:

```{r}
splr <- split(rdat.baysor$cell, rdat.baysor$z_raw)
length(splr)
splr <- lapply(splr, unique)
lengths(splr)
```

Next question will be how to add the coordinates of the segmentation borders 
to the `SpatialExperiment`. Shila Ghazanfour did some work with having instances of
`IntegerList`s in the `spatialCoords` slot [here](https://github.com/shazanfar/spatialStuff/blob/master/reports/SpatialExperiment_polygons_example.Rmd).

5. Clustering / cell type assignment: assignment of each cell to cell type clusters.

```{r, message = FALSE}
ctype.file <- file.path(baysor.dir, "clustering", "cell_assignment.csv")
ctype.baysor <- vroom::vroom(ctype.file)
ctype.baysor <- data.frame(ctype.baysor)
dim(ctype.baysor)
head(ctype.baysor)
table(ctype.baysor$leiden_final)
```

combine with cell metadata

```{r}
stopifnot(all(ctype.baysor$cell == cdat.baysor$cell))
cdat.baysor <- merge(cdat.baysor, ctype.baysor, by = "cell")
head(cdat.baysor)
```

6. Marker genes: Marker gene statistics for each cluster

```{r, message = FALSE}
marker.file <- file.path(baysor.dir, "clustering", "marker_genes.csv")
marker.baysor <- vroom::vroom(marker.file)
marker.baysor <- data.frame(marker.baysor)
dim(marker.baysor)
head(marker.baysor)
```

Quick sanity check whether all markers are present in the count matrix:

```{r, height = 6}
all(marker.baysor$gene_name %in% rownames(counts.baysor))
hist(table(marker.baysor$gene_name))
```

7. Construct `SpatialExperiment`:

Construct `data.frame` of molecule coordinates. Here, we only include the x-, y-, z-coordinates,
but we could keep all columns from the molecule metadata file such as `qc_score` and 
`assignment confidence`.

```{r}
genes <- rdat.baysor$gene
cells <- rdat.baysor$cell
mol <- BumpyMatrix::splitAsBumpyMatrix(rdat.baysor[, c("x", "y", "z")], 
                                       row = genes, col = cells)
```

Quick sanity check for molecules of a specific gene in a specific cell:

```{r}
subset(rdat.baysor, gene == "Neat1" & cell == 1)
mol["Neat1",1]
counts.baysor["Neat1",1]
```

```{r}
stopifnot(all(rownames(mol) == rownames(counts.baysor)))
spe <- SpatialExperiment(assays = list(counts = counts.baysor, molecules = mol),
                         colData = cdat.baysor,
                         spatialCoordsNames = c("x", "y"),
                         spatialDataNames = c("density", "elongation", "area", "avg_confidence"))
spe
```

Add the images:

```{r}
spe <- SpatialExperiment::addImg(spe, 
                                 sample_id = "sample01", 
                                 image_id = "dapi",
                                 imageSource = dapi.file, 
                                 scaleFactor = NA_real_, 
                                 load = FALSE)
spe <- SpatialExperiment::addImg(spe, 
                                 sample_id = "sample01", 
                                 image_id = "membrane",
                                 imageSource = mem.file, 
                                 scaleFactor = NA_real_, 
                                 load = FALSE)
spe
```

7. Inspect `SpatialExperiment`:

```{r}
assay(spe, "counts")[1:5,1:5]
assay(spe, "molecules")["Neat1",1]
colData(spe)
head(spatialCoords(spe))
spatialData(spe)
imgData(spe)
imgData(spe)$data
```

### Cellpose (membrane or DAPI-based) segmentation

Ok, here is a question: where do we store alternative segmentations?
The `altExps` slot of a `SummarizedExperiment`-derivative is thought for alternative
experiments over a distinct set of features for the same samples. But here we are
having a distinct set of samples/cells for the sample features. 
Is this just a separate `SpatialExperiment` with potentially duplicating information
(incl. the images!) or a use case for `MultiAssayExperiment`? 
Or maybe alternative BumpyMatrices in a separate slot such as the `reducedDim` slot.

1. Cell-by-gene matrix. Rows: genes; Cols: cells

```{r, message = FALSE}
counts.file <- file.path(cellpose.dir, "segmentation", "segmentation_counts.tsv")
counts.cellpose <- vroom::vroom(counts.file) 
counts.cellpose <- data.frame(counts.cellpose) 
dim(counts.cellpose)
counts.cellpose[1:5,1:5]
```

2. Cell coordinates:

```{r, message = FALSE}
cdat.file <- file.path(cellpose.dir, "segmentation", "cell_coords.csv")
cdat.cellpose <- vroom::vroom(cdat.file)
cdat.cellpose <- data.frame(cdat.cellpose)
dim(cdat.cellpose)
head(cdat.cellpose)
```

### Baysor (with Cellpose membrane segmentation prior) segmentation

1. Cell-by-gene matrix. Rows: genes; Cols: cells

```{r}
counts.file <- file.path(baysor.cellpose.dir, "segmentation", "segmentation_counts.tsv")
counts.baysor.cellpose <- read.delim(counts.file, header = FALSE) 
dim(counts.baysor.cellpose)
counts.baysor.cellpose[1:5,1:5]
```

2. Cell metadata. Each cell corresponds to each column in `segmentation_counts.csv`.

```{r, message = FALSE}
cdat.file <- file.path(baysor.dir, "segmentation", "segmentation_cell_stats.csv")
cdat.baysor.cellpose <- vroom::vroom(cdat.file)
cdat.baysor.cellpose <- data.frame(cdat.baysor.cellpose)
colnames(counts.baysor.cellpose) <- cdat.baysor.cellpose$cell
dim(cdat.baysor.cellpose)
head(cdat.baysor.cellpose)
```

3. Segmentation and mRNA metadata. Each row corresponds to one mRNA in `raw_data/molecules.csv`

```{r, message = FALSE}
rdat.file <- file.path(baysor.dir, "segmentation", "segmentation.csv")
rdat.baysor.cellpose <- vroom::vroom(rdat.file)
rdat.baysor.cellpose <- data.frame(rdat.baysor.cellpose)
rdat.baysor.cellpose <- subset(rdat.baysor.cellpose, cell != 0)
dim(rdat.baysor.cellpose)
head(rdat.baysor.cellpose)
```   

# Visualization

## Cell metadata

Overlay cell type annotation as in Figure 6 of the publication.

```{r, fig.height = 10}
plot(SpatialExperiment::imgRaster(mem.img))
endo.ind <- spe$leiden_final == "Endothelial"
points(x = spatialCoords(spe)[endo.ind, "x"], 
       y = spatialCoords(spe)[endo.ind, "y"],
       col = "lightgreen", pch = 20)
bplasm.ind <- spe$leiden_final == "B (Plasma)"
points(x = spatialCoords(spe)[bplasm.ind, "x"],
       y = spatialCoords(spe)[bplasm.ind, "y"],
       col = "firebrick", pch = 20)
```

## Segmentation

Let's plot segmentation borders for the first z-layer:

```{r, fig.height = 7}
df1 <- subset(df, z == 1)
ggplot(df1, aes(x = x, y = y)) + 
    geom_polygon(aes(group = cell)) +
    theme_void() 
```

Add holes:

```{r}
.f <- function(df) 
{
    df$x <- df$x + 0.5 * (mean(df$x) - df$x)
    df$y <- df$y + 0.5 * (mean(df$y) - df$y)
    return(df)
}
spl <- split(df1, df1$cell)
dl <- lapply(spl, .f)
holes <- do.call(rbind, dl)  
df1$subid <- 1L
holes$subid <- 2L
df1 <- rbind(df1, holes)
```

Plot with holes:

```{r, fig.height = 8}
ggplot(df1, aes(x = x, y = y)) + 
    geom_polygon(aes(group = cell, subgroup = subid), fill = "firebrick") +
    theme_void() 
```

Plot over image:

```{r, fig.height = 8}
grb <- grid::rasterGrob(mem.img,
    interpolate = FALSE,
    width = unit(1, "npc"),
    height = unit(1, "npc"))
p <- ggplot() + 
    annotation_custom(
        grob = grb,
        xmin = 0,
        xmax = ncol(grb$raster),
        ymin = 0,
        ymax = nrow(grb$raster)) + 
    coord_fixed(
        xlim = c(0, ncol(grb$raster)),
        ylim = c(0, nrow(grb$raster))) 
p <- p + geom_polygon(
            data = df1,
            aes(x = x, y = y, group = cell, subgroup = subid), 
            fill = "lightblue")
p + theme_void()
```

# Interactive exploration with iSEE and Vitessce

Create and process `SingleCellExperiment` for iSEE:

```{r}
sce <- SingleCellExperiment(assays = list(counts = counts.baysor), 
                            colData = cdat.baysor)
sce <- scater::logNormCounts(sce)
sce <- scater::runPCA(sce)
sce <- scater::runTSNE(sce, dimred = "PCA", perplexity = 30)
sce <- scater::runUMAP(sce, dimred = "PCA")
```

Some cosmetics for visualization purposes:

```{r}
sce.sub <- subset(sce, , leiden_final != "Removed")
lev <- levels(factor(sce.sub$leiden_final))
levi <- setdiff(lev, lev[c(1,4,6)])
sce.sub <- subset(sce, , leiden_final %in% levi)
for(col in c("x", "y", "sizeFactor"))
    sce.sub[[col]] <- round(sce.sub[[col]], digits = 3)
sce.sub
```
