---
title: "Visualization of MERFISH mouse ileum data"
author: "Ludwig Geistlinger and Robert Gentleman"
affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse ileum visualization}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r, message = FALSE}
library(cytomapper)
library(vroom)
library(HDF5Array)
```

Start with downloading from and unzipping the dataset from:
https://datadryad.org/stash/dataset/doi:10.5061%2Fdryad.jm63xsjb2

```{r}
data.dir <- "data_release_baysor_merfish_gut"
raw.dir <- file.path(data.dir, "raw_data")
proc.dir <- file.path(data.dir, "data_analysis")
baysor.dir <- file.path(proc.dir, "baysor") 
cellpose.dir <- file.path(proc.dir, "cellpose") 
baysor.cellpose.dir <- file.path(proc.dir, "baysor_membrane_prior") 
```

# Data

Def. ileum: the final and longest segment of the small intestine.

## Raw data

What's there:

```{r}
list.files(raw.dir)
```

mRNA molecule data: 820k observations for 241 genes

```{r, message = FALSE}
mol.file <- file.path(raw.dir, "molecules.csv")
mol.dat <- vroom::vroom(mol.file)
mol.dat <- data.frame(mol.dat)
dim(mol.dat)
head(mol.dat)
length(unique(mol.dat$gene))
```

Image data: 

1. [DAPI](https://en.wikipedia.org/wiki/DAPI) stain signal:

```{r}
dapi.file <- file.path(raw.dir, "dapi_stack.tif")
dapi.img <- EBImage::readImage(dapi.file)
dapi.img
```

2. Membrane Na+/K+ - ATPase immunofluorescence signal:

```{r, fig.height = 10}
mem.file <- file.path(raw.dir, "membrane_stack.tif")
mem.img <- EBImage::readImage(mem.file)
mem.img
```

Store in a `CytoImageList` object:

```{r}
cil <- cytomapper::CytoImageList(dapi = dapi.img, membrane = mem.img)
cil
channelNames
```

# Visualization

Plot pixels with images load in memory, takes relatively long:

```{r, fig.height = 10}
system.time(
    cytomapper::plotPixels(cil)
)
```

This plots the first z-layer/channel, but we can select a specific z-layer/channel of
the 9 different channels in the z-stack tiffs:

```{r, fig.height = 10}
cytomapper::channelNames(cil)
cytomapper::plotPixels(cil, colour_by = "2")
cytomapper::plotPixels(cil, colour_by = "3")
```

## On disk storage of images

```{r}
cur_dir <- HDF5Array::getHDF5DumpDir()
```

Write to HDF5, takes very long:

```{r}
system.time(
    cli.hdf5 <- cytomapper::loadImages(raw.dir, 
                                       pattern = ".tif$",
                                       on_disk = TRUE,
                                       h5FilesPath = cur_dir)
)
cli.hdf5
```

Plot pixels using hdf5 backend: 

```{r, fig.height = 10}
system.time(
    cytomapper::plotPixels(cli.hdf5)
)
```
