---
title: "Import and representation of MERFISH mouse hypothalamus data"
author: "Ludwig Geistlinger and Robert Gentleman"
affiliation: Center for Computational Biomedicine, Harvard Medical School
output:
  BiocStyle::html_document:
    self_contained: yes 
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
vignette: >
  % \VignetteIndexEntry{Mouse hypothalamus}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```

# Setup

```{r, message = FALSE}
library(vroom)
library(SpatialExperiment)
library(ggpubr)
```

Start with downloading from and unzipping the dataset from:
https://datadryad.org/stash/dataset/doi:10.5061/dryad.8t8s248

```{r}
data.file <- "Moffitt_and_Bambah-Mukku_et_al_merfish_all_cells.csv"
```

# Data

Def. hypothalamic preoptic region: is a part of the anterior hypothalamus that 
controls essential social behaviors and homeostatic functions.

Dimensions: 1,027,848 cells; 161 genes

Cell segmentation based on total polyadenylated mRNA and DAPI nuclei costains 
(where are the images?)

Genes: the original publication says:
Combinatorial smFISH imaging was used to identify 135 genes, followed by sequential
rounds of two-color FISH to identify 20 additional genes (but here we have 161 genes?) 

```{r, message = FALSE}
dat <- vroom::vroom(data.file)
dat <- data.frame(dat)
dim(dat)
dat[1:5,1:5]
ind <- grep("Neuron_cluster_ID", colnames(dat))
genes <- colnames(dat)[(ind + 1):ncol(dat)]
genes
length(genes)
```

Note: we also have complementary scRNA-seq of ~31,000 cells dissociated and
captured from the preoptic region of the hypothalamus from multiple male and female mice
available on GEO ([GSE113576](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113576)).

1. Cell-by-gene matrix. Rows: genes; Cols: cells

```{r, message = FALSE}
exprs <- t(as.matrix(dat[,genes]))
dim(exprs)
exprs[1:5,1:5]
```

2. Cell metadata. 

```{r, message = FALSE}
cdat <- dat[,seq_len(ind)]
head(cdat)
```

Data has been obtained from 36 mice (16 female, 20 male)

```{r}
table(cdat$Animal_ID, cdat$Animal_sex)
```

Animal behavior by sex:

```{r}
table(cdat$Behavior, cdat$Animal_sex)
```

Cell type assignment:

```{r}
table(cdat$Cell_class)
```

Def. Bregma: The bregma is the anatomical point on the skull at which the coronal
suture is intersected perpendicularly by the sagittal suture. Used here as a 
reference point for the twelve 1.8- by 1.8-mm imaged slices along the z-axis.

The anterior position of the preoptic region is at Bregma +0.26.

```{r}
table(cdat$Bregma)
```

Some cosmetics on the column names

```{r}
colnames(cdat)[c(2:3,5:7)] <- c("sample_id", "sex", "z", "x", "y")
colnames(cdat) <- tolower(colnames(cdat))
```

It figures that we need to center the data for each bregma slice:

```{r}
sids <- unique(cdat$sample_id)
for(i in sids)
{
    ind <- cdat$sample_id == i
    ccdat <- cdat[ind,]
    spl <- split(ccdat, ccdat$z)
    
    for(i in seq_along(spl)) 
    {   
        spl[[i]]$x <- scale(spl[[i]]$x, scale = FALSE)
        spl[[i]]$y <- scale(spl[[i]]$y, scale = FALSE)
    }

    cdat[ind,] <- do.call(rbind, spl)
}
```

3. Construct `SpatialExperiment`:

```{r}
spe <- SpatialExperiment(assays = list(exprs = exprs),
                         colData = cdat,
                         spatialCoordsNames = c("x", "y", "z"))
spe
```

4. Inspect `SpatialExperiment`:

```{r}
assay(spe)[1:5,1:5]
colData(spe)
head(spatialCoords(spe))
```

5. Visualize as in Figure 3E of the paper for six different anterior-posterior 
   positions from a single female mouse.

```{r, fig.wide = TRUE, fig.width = 10, fig.height = 10}
cdat <- subset(cdat, cell_class != "Ambiguous")
cdat$cell_class <- sub(" [1-4]$", "", cdat$cell_class)
cdat1 <- subset(cdat, sample_id == 1)
cdat1 <- subset(cdat1, z %in% c(0.26, 0.16, 0.06, -0.04, -0.14, -0.24))
cdat1$z <- as.character(cdat1$z)
zum <- paste(0:5 * 100, "um")
names(zum) <- as.character(c(0.26, 0.16, 0.06, -0.04, -0.14, -0.24))
cdat1$z <- unname(zum[cdat1$z]) 

pal <- ggpubr::get_palette("simpsons", 9)
names(pal) <- c("Endothelial", "Excitatory", "OD Immature", "Astrocyte", "Mural", 
                "Microglia", "Ependymal", "Inhibitory", "OD Mature")

ggpubr::ggscatter(cdat1, x = "x", y = "y", color = "cell_class", facet.by = "z",
                  shape = 20, size = 1, palette = pal) +
                  guides(color = guide_legend(override.aes = list(size = 3)))
```


